### 2014-11-14 paper/intro.tex
redid intro

\section*{Types}

Values in Patina can be described by the following types:
\[
\begin{array}{lccl}
\textrm{Lifetime} & \lt & & \\
\textrm{Type Variable} & X & & \\
\textrm{Qualifier} & q & \bnfdef & \qimm \bnfalt \qmut \\
\textrm{Type} & \gt & \bnfdef & \tyint \bnfalt \own{\gt} \bnfalt \tyref{\lt}{q}{\gt} \bnfalt 
				\subvar{\gt}{i}{n} \bnfalt \subrec{\gt}{i}{n} \bnfalt 
				\tyfix{X}{\gt} \bnfalt X \\
\end{array}
\]

>>>>

\section*{Overview}
Rust makes strong claims about ensuring memory safety without garbage collection.
We would like to prove that those claims are true.
To that end, we use a small model of Rust, called Patina, that characterizes the key
features of Rust under the most suspicion, namely its memory management.

Rust memory management has two goals:
no memory should ever be orphaned and no uninitialized memory should ever be read.
Rust tries to achieve this by maintaining two invariants:
all memory has a unique owner responsible for deallocating it and
no memory is ever simultaneously aliased and mutable.
These invariants simplify the situation enough so that
Rust only needs to track which L-values are uninitialized and which have been borrowed.
Ownership tracking of heap memory is managed by unique pointers.
Safe aliasing is managed by borrowed references.

The Patina model has three layers.
The innermost layer deals with L-values and ensures no uninitialized pointer
ever needs to be dereferenced. It also ensures that initialization data correctly
models the runtime memory.
The middle layer deals with R-values.
It ensures that an L-values used do not violate initialization or borrow restrictions.
It also ensures that borrowed references are not created unless they are safe.
The outer layer deals with statements.
It mostly just chains together the guarantees of the L-value and R-value layers.
However, it is also responsible for ensuring no orphans would be created when deallocating memory.

\section*{Types}
Values in Patina can be described by the following types:
\[
\begin{array}{lccl}
\textrm{Lifetime} & \lt & & \\
\textrm{Type Variable} & X & & \\
\textrm{Qualifier} & q & \bnfdef & \qimm \bnfalt \qmut \\
\textrm{Type} & \gt & \bnfdef & \tyint \bnfalt \own{\gt} \bnfalt \tyref{\lt}{q}{\gt} \bnfalt 
				\subvar{\gt}{i}{n} \bnfalt \subrec{\gt}{i}{n} \bnfalt 
				\tyfix{X}{\gt} \bnfalt X \\
\end{array}
\]

### 2014-11-14 paper/paths.tex
redid intro

L-values in Patina are a combination of a variable and a path.
Paths are relative and specify subections of memory reachable from a L-value.

@@

Projection ($\proj{p}{i}$) deconstructs tuples.
Unrolling ($\unroll{\gt}{p}$) deconstructs recursive types.
Dereference ($\deref{p}$) deconstructs pointers and references.
The base path ($\base$) does not deconstruct anything.

@@

With path typing and evaluation now defined, we would like to prove progress
and preservation for them. Preservation will require we define typing for
runtime data (routes, cells, and layouts), which involves a slight bit of trickery
around variant discriminants and the payload route.

Progress will require a way to guarantee that the necessary \textsc{read}s 
can be completed, i.e. any pointers we need to dereference are in fact initialized
(a property we call \emph{shallow initialization}).
We will accomplish this by using a kind of shadow heap, which will simply track
initialization rather than values.
This will require an analogue of \textsc{read} that extracts the initialization
data for a path, a proof that this operation preserves types 
(which itselfs requires typing for this shadow heap),
a coherence check to ensure the shadow heap models the runtime heap,
and a proof that a shallowly initialized path evaluates to a readable route.

@@

\section*{Read Safety}
We are now in position to prove that our read operation is safe.
That is, if the heap is well formed and a route is has some type,
then we can read some layout at that route and that layout has that same type.

@@

\section*{Tracking Initialization}
To properly type check Patina, we need to know which paths point to initialized memory
and which do not. This is important for ensuring we never use uninitialized memory and
that we correctly free heap memory.
We will accomplish this by creating a shadow heap.
Instead of tracking actual values, we will simply track whether a cell is initialized.
Paths will serve the same function for this shadow heap as routes do for the real heap.

@@

\section*{Checking Shallow Initialization}
We will need an analogue of \textsc{read} for our shadow heap for the
simple purpose of extracting the shadow at a path in the same way we
need to extract the layout at a route from the heap.
However, this analogue also serves as a very useful property checker:
any dereferences in the path are dereferences of initialized memory.
This property, called \emph{shallow initialization} is exactly what we need
for path progress. A shallowly initialized path can successfully dereference
all the pointers necessary to evaluate to a route.
For this reason, we call this operation \textsc{shallow}.

To support dereferencing references, we need to construct a shape from the referenced type.
We do this with the \textsc{init} function.

$$ $$

@@

\section*{Shadow Typing}
We need to ensure that the shadows assigned to variables are consistent with the types
assigned to variables, i.e. we want to avoid situations like
$\gG(x)=\tyint$ and $\gU(x)=\refval{\qimm}{\tyint}$.
We can easily do this by defining a typing judgement for shadows and
defining a well-formedness condition for $\gU$ that ensures it is consistent with $\gG$.
Similar to typing for layouts, $\uninit$ can only have types that other holes can have.
\newline

@@

\subsection*{The INIT Function Preserves Types}
\begin{lem}[\textsc{init} Type Preservation]
  %If $\initShadow{\gt}=\gs$ then $\tc{}{\gs}{\gt}$.
  If $\gt$ is closed, then $\tc{}{\initShadow{\gt}}{\gt}$.
\end{lem}

@@

\subsection*{Shallow Preserves Types}
Just as with \textsc{read}, \textsc{shallow} preserves types.

@@

\section*{Shadow Coherence}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
%We require two auxiliary judgments to check for certain initialization states of layouts.
%The $\textsc{init}(H,l)$ judgment checks that $l$ is \emph{fully initialized}, 
%i.e. $\void$ does not occur in any memory reachable from $l$.
%These layouts are safe to use because we cannot access uninitialized memory from them.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot orphan heap memory by doing so.
\newline

@@

We formally require our heap to match our shadow heap with the following:

@@

\section*{Shadow Preservation}
We need to tie the shadow of a shallowly initialized path
to the layout read from a route when the path evaluated to the route.
This will let us carry guarantees from paths and shadows over to routes and layouts.

@@

\section*{Path Progress}
Finally, we have all we need to proof path progress.
Assuming all our contexts and the heap are well-formed,
then a shallowly initialized, well-typed path can evaluate to some route.
We can then rely to path preservation to show that route has the same type as the path.
Read safety then lets us show that we can read a layout at that route
and that it has the same type.

>>>>

L-values in Patina are a combination of a variable and a path.
Paths are relative and specify subections of memory reachable from a L-value.
Projection ($\proj{p}{i}$) deconstructs tuples.
Unrolling ($\unroll{\gt}{p}$) deconstructs recursive types.
Dereference ($\deref{p}$) deconstructs pointers and references.
The base path ($\base$) does not deconstruct anything.

@@

The L-value layer of Patina has two core operations: evaluating a path
to a memory location and reading the value at that location.
We need to characterize the circumstances under which these operations will progress,
and we need to show that these operations preserve certain information, specifically
type and initialization data. Higher layers of Patina will build on these properties.
We shall start with typing paths and then move into defining evaluation for paths.

@@

Our core operations are now defined,
but we still need to describe the typing our runtime data
so that we can properly state and prove our type preservation lemmas.

%With path typing and evaluation now defined, we would like to prove progress
%and preservation for them. Preservation will require we define typing for
%runtime data (routes, cells, and layouts), which involves a slight bit of trickery
%around variant discriminants and the payload route.

%Progress will require a way to guarantee that the necessary \textsc{read}s 
%can be completed, i.e. any pointers we need to dereference are in fact initialized
%(a property we call \emph{shallow initialization}).
%We will accomplish this by using a kind of shadow heap, which will simply track
%initialization rather than values.
%This will require an analogue of \textsc{read} that extracts the initialization
%data for a path, a proof that this operation preserves types 
%(which itselfs requires typing for this shadow heap),
%a coherence check to ensure the shadow heap models the runtime heap,
%and a proof that a shallowly initialized path evaluates to a readable route.

@@

\section*{Read Safety}
We are now in position to prove that our read operation is safe.
That is, if the heap is well formed and a route is has some type,
then we can read some layout at that route and that layout has that same type.
This is effectively a merger of the type preservation and progress lemmas for \textsc{read}.

@@

\section*{Tracking Initialization}
Types are only half of the information we need to preserve.
To properly type check Patina, we need to know which paths point to initialized memory
and which do not. This is important for ensuring we never use uninitialized memory and
that we correctly free heap memory.
We will accomplish this by creating a shadow heap.
Instead of tracking actual values, we will simply track whether a cell is initialized.
Paths will serve the same function for this shadow heap as routes do for the real heap.

@@

\section*{Checking Shallow Initialization}
We will need an analogue of \textsc{read} for our shadow heap for the
simple purpose of extracting the shadow at a path in the same way we
need to extract the layout at a route from the heap.
However, this analogue also serves as a very useful property checker:
any dereferences in the path are dereferences of initialized memory.
This property, called \emph{shallow initialization} is exactly what we need
for path progress. A shallowly initialized path can successfully dereference
all the pointers necessary to evaluate to a route.
For this reason, we call this operation \textsc{shallow}.
To support dereferencing borrowed references, 
we need to construct a shape from the referenced type.
We do this with the \textsc{init} function.
\newline

@@

\section*{Shadow Typing}
We need to ensure that the shadows assigned to variables are consistent with the types
assigned to variables, i.e. we want to avoid situations like
$\gG(x)=\tyint$ and $\gU(x)=\refval{\qimm}{\tyint}$.
We can easily do this by defining a typing judgement for shadows and
defining a well-formedness condition for $\gU$ that ensures it is consistent with $\gG$.
Similar to typing for layouts, $\uninit$ can only have types that other holes can have.

@@

\subsection*{The INIT Function Preserves Types}
Since the \textsc{init} function is supposed to be the
fully initialized shadow of the argument type, it should
be the case that the resulting shadow has that type.
Since \textsc{init} is only defined on closed types, 
we must restrict ourselves to them.
\begin{lem}[\textsc{init} Type Preservation]
  If $\gt$ is closed, then $\tc{}{\initShadow{\gt}}{\gt}$.
\end{lem}

@@

\subsection*{Shallow Preserves Types}
Just as with \textsc{read}, \textsc{shallow} preserves types.
This lemma is used primarily to demonstrate the impossibility 
of certain subcases of dereferences in later proofs.

@@

\section*{Shadow Coherence}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
%We require two auxiliary judgments to check for certain initialization states of layouts.
%The $\textsc{init}(H,l)$ judgment checks that $l$ is \emph{fully initialized}, 
%i.e. $\void$ does not occur in any memory reachable from $l$.
%These layouts are safe to use because we cannot access uninitialized memory from them.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot orphan heap memory by doing so.
\newline

@@

@@

\section*{Shadow Preservation}
The second main preservation lemma for our core operations.
The initialization data, in the form of a shadow,
is preserved by path evaulation and \textsc{read}
so that the resulting layout is correctly described by the shadow.
%We need to tie the shadow of a shallowly initialized path
%to the layout read from a route when the path evaluated to the route.
%This will let us carry guarantees from paths and shadows over to routes and layouts.

@@

\section*{Path Progress}
Finally, we have all we need to proof path evaluation progress under the right circumstances.
Assuming all our contexts and the heap are well-formed,
then a shallowly initialized, well-typed path can evaluate to some route.
We can then rely to path preservation to show that route has the same type as the path.
Read safety then lets us show that we can read a layout at that route
and that it has the same type. Shadow preservation guarantees us that our
model of the layout reflects reality. All of the key pieces we need for
the higher layers of Patina.
